---
title: A Walk Though of Accessing Financial Statements with XBRL in R - Part 1
author: David Lucey
date: '2020-02-18'
slug: a-walk-though-of-accessing-financial-statements-with-xbrl-in-r-part-1
categories:
  - R
  - XBRL
tags:
  - XBRL
  - rstats
output:
  html_document:
    code_folding: true
---

```{r 'set-up', include=FALSE}

packages <- 
  c("jsonlite",
    "ggplot2",
    "edgarWebR",
    "XBRL",
    "finstr",
    "data.table",
    "stringr")

if (length(setdiff(packages,rownames(installed.packages()))) > 0) {
  install.packages(setdiff(packages, rownames(installed.packages())))  
}

invisible(lapply(packages, library, character.only = TRUE))

knitr::opts_chunk$set(comment=NA, fig.width=12, fig.height=8, out.width = '100%')

```


```{r 'xbrl-specs', include=FALSE}

# XBRL package guidance
options(stringsAsFactors = FALSE)

```


```{r 'load_data', message=FALSE, warning=FALSE, include=FALSE}

# Re-load data previously stored for purposes of this blog post
xbrl_data <- readRDS("/Users/davidlucey/Desktop/David/Projects/xbrl_investment/data/xbrl_lists.RDS")
stmt_list <- xbrl_data$stmt_list
result_list <- xbrl_data$result_list
filing_list <- xbrl_data$filing_list

# Clean up
rm(xbrl_data)

```


# Introduction

As financial professionals and analytic software lovers, the ability to efficiently load a large number of financial statements, and conduct an analysis has always been a key objective. In previous posts, Redwall Analytics worked with a 15-year time series of municipal Comprehensive Annual Financial Reports (CAFR) for 15 Fairfield County, CT towns [Fairfield County Town Level Spending and Liabilities Gallop Since 2001](https://redwallanalytics.com/2018/12/26/fairfield-county-town-level-spending-and-liabilities-gallop-since-2001/). We also studied the worrisome long-term, time series of unfunded liabilities for all 169 Connecticut municipalities [Connecticut City Unfunded Pension And OPEB Liabilities](https://redwallanalytics.com/2019/10/11/connecticut-city-unfunded-pension-and-opeb-liabilities-over-time/). We are probably most proud of our work replicating the complicated one-year (2016) spreadsheet analysis by Marc Joffe (Reason Foundation) and Mark Fitch (Yankee Institute) over the long-term [Replicating Yankee Institute Risk Score Over 15 Years](https://redwallanalytics.com/2019/10/12/replicating-yankee-institute-risk-score-over-15-years/). 

However, all of these were conducted using downloaded .csv data from the State's website, not on the kind of structured XBRL data now required in real-time for public companies on the SEC's Edgar website. Despite several attempts, XBRL had remained elusive up until now, but with daily practice, tasks previously beyond reach, suddenly become achievable. 
A good deal of credit and much of the code to extract XBRL data from Edgar comes Aaron Mumala's 2018 blog post here: [Accessing Financial Data from the SEC - Part 2](https://aaronmams.github.io/Accessing-financial-data-from-the-SEC-Part-2/). In addition, Micah Waldstein's [Parsing Functions in edgarWebR](https://micah.waldste.in/blog/blog/2017/10/parsing-functions-in-edgarwebr/), as well as his edgarWebR package were essential. Lastly, none of this would have been possible without Darko Bergant's [finstr](https://github.com/bergant/finstr) and Roberto Bertolusso's [XBRL](https://cran.r-project.org/web/packages/XBRL/XBRL.pdf) packages. To summarize the workflow to be described below, filing data is located with edgarWeb, the XBRL Instance Documents are downloaded and parsed into a list of data.frames with the XBRL package's xbrlDoAll function, and finally, statements are organized into traditional Income Statements and Balance Sheets using the finstr package. 

This will be the first in a three-part series. First, a quick walk-through to try to fill in some gaps from Aaron Mumala's blog post and current documentation to help others up the learning curve with XBRL. Even better, we would welcome comments clarifying solutions to unresolved issues or even more efficient ways of extracting the same data than we have used. In the second part, we will use another free source of financial statement data (financialmodelingprep.com API) to pull 10-year's of income statements for ~700 publically-listed pharma companies and explore R&D spending. Lastly, we will look to mine financial statement items for warning signs using metrics from a 2003 piece in the CFA Conference Proceeding.


# 10-K Filing List via edgarWebR


Below, we choose to query all past Apple 10-K filings up until last week wrapped in the try() function, so that our lapply of company_filings (from edgarWeb) won't be halted (even if a particular filing can't be successfully loaded). 28 results are available along with the web link (href) to each set of filings. Although filings prior to 2008 are available, these are in html format, and unfortunately, can't be parsed with finstr and XBRL. A later effort may be made to retrieve these by scraping and parsing the html, but this is an advanced operation. 
 

```{r 'filings-list', echo=TRUE, eval=FALSE}

# This chunk was run previously and filing_list was saved for the purposes of this blog post

# Call company_filings from edgarWeb package
filing_list <- 
    try(company_filings(as.character('AAPL'),
                        ownership = FALSE, 
                        type = '10-K', 
                        before = "2020207",
                        count = 40, 
                        page = 1),
        TRUE)

# Show full list
filing_list

```


# List of Annual Index Filing hrefs via edgarWebR


```{r 'href-list', echo=TRUE}

ticker <- 'AAPL'
type <- '10-K'

#parse this listing to find only 10-K filings...e.g. not 10-K/A and 10-K405
href_list <- filing_list[filing_list$type==type,]$href

# Show the first five index links
href_list[1:5]

```


# Individual Filing Package Components via edgarWebR

Here we use edgarWeb's filing_details function to see the available 19 documents for the 2017 index filing package using the 3rd element of the href_list above (2017). The first element here is the html 10-K, and there are links to many other elements from the filing including:  html exhibits, .jpg's, the complete submission as a text file, the XBRL schema and the all important XBRL Instance Document. If we copied the html url 10-K in our browser, it would show the 10-K, however, not in the XBRL form we are looking for. 


```{r 'filing-details', echo=TRUE}


# 2nd Element is 2018
filing_2018 <- 
  filing_details(href_list[[2]])$documents

# Just lines with "XBRL" and "10-K"
filing_2018[str_detect(filing_2018$description,"XBRL|10-K"),]


```


# Extract Available XBRL Instance Documents from Edgar via edgarWebR and XBRL Package Custom Function

In this case, we are going with only 10-K's, so we filter above again by type equal "10-K", and extract the 24 href's available. Next below, we use xbrlDoAll from the XBRL package to extract the financial statements using the links from the href_list we made above. Note that we again use try() because the function would fail on some of the documents. We are not going to run the code here on the blog, because it takes a few minutes to run for Apple. We previously saved and reloaded the object above.

We have not shown here, but XBRL parsing for US companies is governed by the schemas from the US GAAP reported Taxonomy, which are updated annually. These can be found here on the XBRL US website [XBRL Taxonomy](https://xbrl.us/xbrl-taxonomy/2019-us-gaap/). By downloading the relevant .xsd file, removing the .xml suffix at the end and moving it to in the "xbrl.Cache" file created when you run XBRL, these problems were resolved for XBRL Instance Documents. Redwall spent quite a bit of time figuring out that we needed currency, dei, exch, country and currency files among others. We did this manually and by trial and error, but surely there is a better way, but a more efficient solution might be found here on Stack Overflow [XBRL Package Error in File](https://stackoverflow.com/questions/52170323/xbrl-package-error-in-filefromcachefile)


```{r 'stmt-list', echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}

# This chunk was run previously and stmt_list was saved for the purposes of this blog post

# Need to use project where xbrl.cache is stored for this chunk
setwd("/Users/davidlucey/Desktop/David/Projects/xbrl_investment")

# Run xbrlDoAll and store in stmt_list
stmt_list <-  
  
    # apply function to each element of href_list
    lapply(href_list, function(href) {
      
    # Get filing details for the particular annual href index
    filing.details <- filing_details(href)$documents
    
    # Further get XBRL Instance Document for year href
    doc <- filing.details[str_detect(filing.details$description,"XBRL INSTANCE DOCUMENT"),]$href
    
    # Extract XBRL of using specified href
    try(xbrlDoAll(doc))
    })

```


As we have set it up in the code above, we loop through all Apple's 10-K hrefs parsing with the XBRL package. This gives a list object for each annual reporting package, each in turn containing 9-10 data.frames (as shown below). Most of the data.frames relate to the taxonomy, governed by the .xsb files for that year. Starting in 2008, 10-Ks began being filed as "XBRL Instance Documents", and these work seamlessly with the workflow described above. Around 2018, 10-K's shifted to "Extended XBRL Instance Document" format, which we could not build into financial statements with finstr. 


```{r 'show-stmt', echo=TRUE}

# Show one of extracted XBRL statements
summary(stmt_list[[3]])

```

In the future, we would like to better understand the XBRL data structure, but again, it wasn't easy to find much simple documentation on the subject. This post from Darko Bergant gives an excellent diagram of the structural hierarchy of the XBRL object [Exploring XBRL files with R](https://github.com/bergant/XBRLFiles). He goes on in the same post: *All values are kept in the fact table (in the fact field, precisely). The element table defines what are these values (the XBRL concepts, e.g. “assets”, “liabilities”, “net income” etc.). The context table defines the periods and other dimensions for which the values are reported*.

Below, we show the first 50 lines of the "fact" list of the 2017 XBRL package. As noted above, data here is nested in levels, with the top level being the Balance Sheet, Income Statement, etc (again see [Exploring XBRL files with R](https://github.com/bergant/XBRLFiles)). It is possible to drill down to lower level items. Mr. Bergant also explains here how to for example extract a lower level item such as warranty information which wouldn't be broken out at the top level [How to get all the elements contained in the original XBRL?](https://github.com/bergant/finstr/issues/4).


```{r 'xbrl-fact', echo=TRUE}

# Drill down on financial statement "fact" for individual year in stmt_list
stmt_list[[3]]$fact[1:50,c(2:4)]

```

# Apply XBRL_get_statements via finstr Package to transform XBRL into Financial Statements


Next, we run finstr's xbrl_get_statements on our stmt_list objects (ie: the XBRL Instance Documents) to convert the parsed XBRL objects into financial statements.

```{r 'xbrl-get-statements', echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}

# This chunk was run previously and result_list was saved for the purposes of this blog post

result_list <- 
    sapply(stmt_list, function(stmt) {
      try(xbrl_get_statements(stmt))
      })

```


As shown above, finstr's xbrl_get_statements is unsuccessful on the first item (2019), which is the new "Extended XBRL Instance Document", because of a problem with duplicate keys after the spreading the data.frame. It seems that finstr hasn't been updated since 2017, so it likely has something to do with the problem. We asked for guidance on Stack Overflow [Finstr Get XBRL Statement Error Parsing XBRL Instance Documents](https://stackoverflow.com/questions/60152621/finstr-get-xbrl-statement-error-parsing-extracted-xbrl-instance-document), but so far no luck. Also, it fails with the pre-2008 documents as expected, which were only available as html.



# Catching Errors in Successfully Parsed Statements


There are many issues with XBRL including that companies classify items in differing ways, use different names for the same items and sometimes the there are errors in the calculation hierarchy. Classification differences will be challenging, but finstr has the check_statement function to show where there are calculation inconsistencies.


```{r 'check-statement', echo=TRUE}

# Run check_statement on 2nd element of result_list (2018)
for(list in result_list[2]) {
   print(lapply(list, check_statement))
}

```


Above we can see that there are six errors in the "CashAndCashEquivalentsPeriodIncreaseDecrease" element_id in the 2017 Statement of Cash Flows. We can then drill down to see what those are. In this case, it is not a mismatch between the original data and the amount calculated as a check. The check is just not there (NA). We need to do further research to fully understand this function.


```{r}

check <- check_statement(result_list[[2]]$ConsolidatedStatementsOfCashFlows, element_id = "CashAndCashEquivalentsPeriodIncreaseDecrease")

# Calculated is NA
check$calculated

```


# Looking at Apple's 2017 Balance Sheet Disclosure


Here we show the balance sheets for the first available document which is as of September 2018 (2019 does exist, but we were not able to parse that thus far). We could do the same for the Income Statement or the Statement of Cash Flows.


```{r 'balance-sheet', echo=TRUE, message=FALSE, warning=FALSE}

# Drill down to annual balance sheet from result_list
result_list[[2]]$ConsolidatedBalanceSheets

```


Our results_list is nested, so not so easy to work with so we use the purrr package to flatten it into 35 financial statement objects including all of the balance sheets, income statements and cash flows for the period. The same first balance sheet we have already looked at is shown, but now at the top level so that the underlying data is easier to analyze and plot.


```{r 'flatten-financial-statements', echo=TRUE, message=FALSE, warning=FALSE}

# Flatten nested lists down to one list of all financial statements for the 10 year period
fs <- purrr::flatten(result_list)

# Now the balance sheet is at the top level of fs instead of nested
fs[[2]]

```


Next, we can graph the evolution of "Current" items over the 10-year period. Note that in 2017, Apple started calling them "Consolidated Balance Sheets". Prior to that, they were named "Statements of Financial Position Classified". Hence, we had to regex match using both "Balance" and "Position" to get the relevant documents for the full period. This clearly cumbersome and difficult to do at scale on a larger number of companies. We can see that both current assets and liabilities have tripled over the period.


```{r 'plot-current', echo=TRUE, message=FALSE, warning=FALSE}
# Regex match list items matching "Balance" and "Position" and rbind into bs (balance sheet)
bs <-
  rbindlist(fs[str_detect(names(fs), "Balance|Position")], fill = TRUE)

# Drop any duplicates with same endDate
bs <- unique(bs, by = "endDate")

# Function to scale y axis label
scaleFUN <- function(x) paste("$",x/1000000000,"Billion")

current <- names(bs)[str_detect(names(bs), "Current")]

# Tidy and ggplot from within data.table bs object coloring variables
bs[, melt(.SD, id.vars = "endDate", measure.vars=current)][
  ][!is.na(value)][
    ][, ggplot(.SD, 
               aes(as.Date(endDate),
                   as.numeric(value),
                   color = variable)) +
        geom_line() +
        scale_y_continuous(labels = scaleFUN) +
        labs(title = "Apple Current Items over Time",
             caption = "Source: SEC Edgar") +
        ylab("Amount") +
        xlab("Year") +
        theme_bw()]

```


We are curious about R&D spending over time, so here we took the income statements which again used changing names (Statements of Income and Statements of Operations) over time. It looks like Apple reduced its R&D spending sharply after the iPhone launch, but has ramped it up again to launch the wearables business. We will use this strategy to extract specific balance sheet and income statement items in Part 3 to mine for warning signs.


```{r}
# Regex match list items matching "StatementOFIncome" and StatementsOfOperations", and rbind into is (income statement)
is <-
  rbindlist(fs[str_detect(names(fs),"StatementOfIncome$|StatementsOfOperations$")], fill =
              TRUE)

# Drop NA rows
is <- 
  is[!is.na(SalesRevenueNet)]

# Drop previous year's which are dupes
is <-
  unique(is, by = "endDate")

# Mutate R&D to sales ratio variable and select columns needed
is <- 
  is[, rd_sales :=
       ResearchAndDevelopmentExpense / SalesRevenueNet][
       ][, .(endDate, rd_sales)]

# Tidy and ggplot within data.table is object
is[, melt(.SD, id.vars = "endDate", measure.vars=c("rd_sales"))][
  , ggplot(.SD, aes(as.Date(endDate), as.numeric(value), color = variable)) +
    geom_line() +
    scale_y_continuous(labels = scales::percent) +
    labs(title = "Apple R&D-to-Sales Ratio has More than Doubled since 2012",
         caption = "SEC Edgar") +
    ylab("Amount") +
    xlab("Year") +
    theme_bw()]

```

# Conclusion

After spending the few years solving many problems in R, the volume of discussion about XBRL still seems surprisingly sparse for such a big potential use case. In addition, most of development at least in R seemed to stop cold in 2017. It will be interesting to learn if this is because most the people who look at financial statements generally don't use XBRL in analytic software yet, or because it is too inefficient to get clean data in a usable form. In the end, Aaron Mumala suggested that downloading and parsing XBRL from Edgar was probably still not ready for prime time.

Admittedly, the challenges discovered in this exercise because of changing statement names, financial statement items, data formats, we wondered if the SEC Edgar XBRL disclosures are suitable for a large scale analysis. Many providers charge for data parsing and cleaning, and it may be worth paying for, especially if real money will be involved. In Part 2 of this series, we will look at using a free outside provider of financial statement data (Financial Modeling Prep) and see what we find.



```{r 'save-rds', eval=FALSE, include=FALSE}
# Save
saveRDS(object=list(stmt_list=stmt_list,result_list=result_list,filing_list=filing_list),"xbrl_lists.RDS")
```


